---
title: "Clean_Code"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries needed:


```{r}
library(tidyverse)
library(dplyr)
library(funModeling) 
```

## Loading Hotel Bookings dataset:

```{r}
hotels_df = read.csv('hotel_bookings.csv')
```

## Initial data cleanup:

* Changing the reservation status date column to date type:

```{r}
hotels_df$reservation_status_date = as.Date(hotels_df$reservation_status_date, format="%Y-%m-%d")
```

* Changing all character type data columns to factor types:

```{r}
hotels_df = hotels_df %>% mutate_if(is.character, as.factor)
```

* Dropping null values:

```{r}
hotels_df = hotels_df %>% drop_na(children)
```

* Looking at number cancelled bookings versus not for each hotel type:

```{r}
cancel_df = hotels_df %>% select(hotel, is_canceled) %>%
                group_by(hotel, is_canceled) %>%
                summarise(cancellation_type_count = n()) %>%
                mutate(total_bookings = sum(cancellation_type_count)) %>%
                mutate(percent_cancellation
                     = round((cancellation_type_count/total_bookings)*100, 2)) 
```

* Changing binary values to strings for better visual understanding:

```{r}
cancel_df$is_canceled = ifelse(cancel_df$is_canceled == 0,'Not Cancelled','Cancelled')
```

```{r}
cancel_df = cancel_df %>% mutate_if(is.character, as.factor)
```


```{r}
repeat_df = hotels_df %>% select(hotel, is_canceled, is_repeated_guest,
              stays_in_weekend_nights, stays_in_week_nights) %>%
              filter(is_canceled == 1) %>%
              group_by(hotel, is_canceled, is_repeated_guest)%>%
              summarise(bookings_by_guest_type = n(),
              weekend_bookings = sum(stays_in_weekend_nights),
              weekday_bookings = sum(stays_in_week_nights))
```

* Changing binary values to strings for better visual understanding:

```{r}
repeat_df$is_repeated_guest = ifelse(repeat_df$is_repeated_guest == 0,
                                     'Not Repeated guest','Repeated Guest')
```

```{r}
repeat_df = repeat_df %>% mutate_if(is.character, as.factor)
```

```{r}
cancel_df
```

* Write data csv files for app use:

```{r}
# write.csv(cancel_df, file='cancel_df.csv', row.names=FALSE)
```

```{r}
# copy_cancel_df = read.csv('cancel_df.csv', stringsAsFactors = TRUE)
```

```{r}
# copy_cancel_df
```

```{r}
# repeat_df
```

```{r}
# write.csv(repeat_df, file='repeat_df.csv', row.names=FALSE)
```

```{r}
# copy_repeat_df = read.csv('repeat_df.csv', stringsAsFactors = TRUE)
```

```{r}
# copy_repeat_df
```

```{r}
hotels_df
```

```{r}
write.csv(hotels_df, file='hotels_df.csv', row.names=FALSE)
```

```{r}
getwd()
```


```{r}
copy_hotels_df = read.csv('hotels_df.csv', stringsAsFactors = TRUE)
```

```{r}
copy_hotels_df
```



