---
title: "Hotel_Bookings"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries needed:

```{r}
# install.packages("treemap", repos="http://cran.rstudio.com/", dependencies=TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(funModeling) 
library(treemap)
library(RColorBrewer)
```

## Loading Hotel Bookings dataset:

```{r}
hotels_df = read.csv('hotel_bookings.csv')
```

## Initial data cleanup:

* Changing the reservation status date column to date type:

```{r}
hotels_df$reservation_status_date = as.Date(hotels_df$reservation_status_date, format="%Y-%m-%d")
```

* Changing all character type data columns to factor types:

```{r}
hotels_df = hotels_df %>% mutate_if(is.character, as.factor) %>%
                          glimpse()
```

```{r}
dim(hotels_df)
```

```{r}
head(hotels_df, 6)
```

## Dataset makeup:

```{r}
str(hotels_df)
describe(hotels_df)
```

## Checking null values:

* Checking total sum of null values:

```{r}
sum(is.na(hotels_df))
```

* Checking which columns these come from:

```{r}
df <- as.data.frame(
  cbind(
    lapply(
      lapply(hotels_df, is.na), sum)
    )
  )

rownames(subset(df, df$V1 != 0))
```

* Dropping null values:

```{r}
hotels_df = hotels_df %>% drop_na(children)
```

```{r}
dim(hotels_df)
```

```{r}
hotels_df
```


## Looking at cancellations for each hotel type:

### Overall correlation and cancellation correlation:

```{r}
correlation_table(hotels_df, 'is_canceled')
```

```{r}
# cor(hotels_df %>% filter(is_canceled == 0) %>% select(is_canceled, lead_time, arrival_date_year, arrival_date_week_number, arrival_date_day_of_month, stays_in_weekend_nights, stays_in_week_nights, adults, children, babies, is_repeated_guest, previous_cancellations, previous_bookings_not_canceled, booking_changes, booking_changes, days_in_waiting_list, required_car_parking_spaces, required_car_parking_spaces, total_of_special_requests))
```

* Looking at number of bookings in the dataset for each hotel type:

```{r}
freq(hotels_df['hotel'], plot = FALSE)
```
* Looking at number cancelled bookings versus not for each hotel type:

```{r}
cancel_df = hotels_df %>% select(hotel, is_canceled) %>%
                group_by(hotel, is_canceled) %>%
                summarise(cancellation_type_count = n()) %>%
                mutate(total_bookings = sum(cancellation_type_count)) %>%
                mutate(percent_cancellation
                     = round((cancellation_type_count/total_bookings)*100, 2)) 
```

* Changing binary values to strings for better visual understanding:

```{r}
cancel_df$is_canceled = ifelse(cancel_df$is_canceled == 0,'Not Cancelled','Cancelled')
```

```{r}
cancel_df = cancel_df %>% mutate_if(is.character, as.factor)
```

```{r}
cancel_df
```

```{r}
cancel_df %>% select(hotel, is_canceled, percent_cancellation) %>%
              ggplot(. ,aes(x = hotel, 
                            y = percent_cancellation, 
                            label = percent_cancellation,
                            fill = is_canceled)) + 
              geom_col(stat = "identity", width = 0.7) + 
              geom_text(size = 3.5, position = position_stack(vjust = 0.5), 
                        colour = "white", fontface = "bold") +
              labs(title ="Percent Cancellations by Hotel Type", x = "Type of Hotel",
                       y = "Percentage of bookings (%)") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())
```
```{r}
cancel_df %>% select(hotel, is_canceled, percent_cancellation) %>%
              filter(hotel == "City Hotel") %>%
              ggplot(. ,aes(x = hotel, 
                            y = percent_cancellation, 
                            label = percent_cancellation,
                            fill = is_canceled)) + 
              geom_col(stat = "identity", width = 0.5) + 
              geom_text(size = 3.5, position = position_stack(vjust = 0.5), 
                        colour = "white", fontface = "bold") +
              labs(title ="Percent Cancellations by Hotel Type", x = "Type of Hotel",
                       y = "Percentage of bookings (%)") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())

```
```{r}
cancel_df %>% select(hotel, is_canceled, percent_cancellation) %>%
              filter(hotel == "Resort Hotel") %>%
              ggplot(. ,aes(x = hotel, 
                            y = percent_cancellation, 
                            label = percent_cancellation,
                            fill = is_canceled)) + 
              geom_col(stat = "identity", width = 0.5) + 
              geom_text(size = 3.5, position = position_stack(vjust = 0.5), 
                        colour = "white", fontface = "bold") +
              labs(title ="Percent Cancellations by Hotel Type", x = "Type of Hotel",
                       y = "Percentage of bookings (%)") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())

```

```{text}

* For the city hotel it seems like we have a 60:40 split whereas for the Resort hotel we have a 72:28 split. The resort hotel looks better in terms of the cancelled vs not-cancelled ratios. We can focus more on City Hotel for further cancellation analysis.

```

## Looking at number of bookings per deposit type for each hotel type:

* Deposit types:

    + No Deposit – no deposit was made
    + Non Refund – a deposit was made in the value of the total stay cost
    + Refundable – a deposit was made with a value under the total cost of stay

```{r}
hotels_df %>% select(hotel, is_canceled, deposit_type) %>%
              group_by(hotel, is_canceled, deposit_type) %>%
              summarise(bookings_by_deposit_type = n())
```

```{r}
is_canceled_labs = c("Not cancelled", "Cancelled")
names(is_canceled_labs) <- c("0", "1")

hotels_df %>% select(hotel, is_canceled, deposit_type) %>%
           group_by(hotel, is_canceled, deposit_type) %>%
           summarise(bookings_by_deposit_type = n()) %>%
           ggplot(. ,aes(x = deposit_type, y = log(bookings_by_deposit_type), label = bookings_by_deposit_type)) +
           geom_col(aes(fill = hotel), position = "dodge2") +
           facet_grid(is_canceled ~ hotel, labeller = labeller(is_canceled = is_canceled_labs)) + 
           scale_fill_manual(values=c("deeppink4","darkorchid4")) +
           labs(title = "Number of bookings for each deposit type",
                x = "Type of Deposit",
                y = "Number of bookings - log scale") +
           geom_text(size = 3.5, position = position_dodge2(0.9), vjust = 1.5, 
                        colour = "white", fontface = "bold") +
           theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank()) 
```
```{r}
is_canceled_labs = c("Not cancelled", "Cancelled")
names(is_canceled_labs) <- c("0", "1")

hotels_df %>% select(hotel, is_canceled, deposit_type) %>%
           filter(is_canceled == 0) %>% 
           group_by(hotel, is_canceled, deposit_type) %>%
           summarise(bookings_by_deposit_type = n()) %>%
           ggplot(. ,aes(x = deposit_type, y = log(bookings_by_deposit_type), label = bookings_by_deposit_type)) +
           geom_col(aes(fill = hotel), position = "dodge2") +
           facet_grid(is_canceled ~ hotel, labeller = labeller(is_canceled = is_canceled_labs)) + 
           scale_fill_manual(values=c("deeppink4","darkorchid4")) +
           labs(title = "Number of bookings for each deposit type",
                x = "Type of Deposit",
                y = "Number of bookings - log scale") +
           geom_text(size = 3.5, position = position_dodge2(0.9), vjust = 1.8, 
                        colour = "white", fontface = "bold") +
           theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank()) 

```
```{r}

hotels_df %>% select(hotel, is_canceled, deposit_type) %>%
           filter(is_canceled == 1) %>% 
           group_by(hotel, is_canceled, deposit_type) %>%
           summarise(bookings_by_deposit_type = n()) %>%
           ggplot(. ,aes(x = deposit_type, y = log(bookings_by_deposit_type), label = bookings_by_deposit_type)) +
           geom_col(aes(fill = hotel), position = "dodge2") +
           facet_grid(is_canceled ~ hotel, labeller = labeller(is_canceled = is_canceled_labs)) + 
           scale_fill_manual(values=c("deeppink4","darkorchid4")) +
           labs(title = "Number of bookings for each deposit type",
                x = "Type of Deposit",
                y = "Number of bookings - log scale") +
            +
           theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank()) 

```

```{text}

Upper half of the first graph - non-cancellations for both hotels = For the bookings that are not cancelled and go on to successfully generate revenue, majority of the bookings paid no deposit. So majority of customers, who are not intending to stay don't care about the deposit amount. However, if in this scenario the customers decided to cancel en-masse, most of these cancellations would come from the non-deposit category as those number are quite higher than the deposit paid categories.

Lower half of the first graph - cancellations for both hotels = The total number of bookings cancelled after paying no deposit is higher than people cancelling after paying some deposit. The hotel's cancelled numbers could be lower if some deposit is taken.
```

## Looking at distribution types and average daily rates for both hotel types:

* Distribution types:

    + Direct was assumed to be direct hotel booking, actual key not provided
    + TA - Travel Agents , TO - Tour Operators
    + No explanation for Corporate and GDS.
    
* Number of bookings per distribution type:

```{r}
hotels_df %>% select(hotel, is_canceled, distribution_channel) %>%
              filter(is_canceled == 1) %>%
              group_by(hotel,is_canceled, distribution_channel) %>%
              summarise(bookings_by_distribution = n()) %>%
              arrange(bookings_by_distribution)

```

```{r}

hotels_df %>% select(hotel, is_canceled, distribution_channel) %>%
              filter(is_canceled == 0) %>%
              group_by(hotel,is_canceled, distribution_channel) %>%
              summarise(bookings_by_distribution = n()) %>%
              arrange(bookings_by_distribution)

```

* Focusing on Direct and TA/TA (cancelled) as they had majority of the bookings:

* Cancelled bookings by distribution type for both hotel types:

```{r}
hotels_df %>% select(hotel, is_canceled, distribution_channel) %>%
              filter(is_canceled == 1) %>%
              group_by(hotel, is_canceled, distribution_channel) %>%
              summarise(bookings_by_distribution = n()) %>%
              mutate(total_cancelled_bookings = sum(bookings_by_distribution)) %>%
              ggplot(. ,aes(x = log(bookings_by_distribution), y = hotel,
                            label = bookings_by_distribution, fill = distribution_channel)) +
              geom_col(stat = "identity", width = 0.5) +
              geom_text(size = 3.5, position = position_stack(vjust = 0.5), 
              colour = "white", fontface = "bold") +
              labs(title ="Number of cancelled bookings per distribution type", y = "Type of Hotel",
                       x = "Number of bookings - (log-scale)") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())
```

* Average ADR for cancelled bookings by distribution type for both hotel types:

```{r}
hotels_df %>% select(hotel, is_canceled, distribution_channel, adr) %>%
              filter(is_canceled == 1 & adr<400) %>%
              group_by(hotel, is_canceled, distribution_channel) %>%
              summarise(avg_adr = mean(adr))
```

```{text}

In cancelled bookings, TA/TO bookings formed the majority over other types - the reason probably is easy of cancellation and low rates that customer sees in this case while cancelling. For the City hotel, the direct booking rate is slightly higher than the TA/TO rate, not the same again for Resort hotel.  Lower Direct rate to match TA/TO rate + plus take deposit to possible reduce a bit of cancellations.

```

* Non-cancelled bookings by distribution type for both hotel types:

```{r}
hotels_df %>% select(hotel, is_canceled, distribution_channel) %>%
              filter(is_canceled == 0) %>%
              group_by(hotel,is_canceled, distribution_channel) %>%
              summarise(bookings_by_distribution = n()) %>%
              mutate(total_non_cancelled_bookings = sum(bookings_by_distribution))
```

```{text}

In non-cancelled bookings, show the same trend - TA/TO bookings formed the majority over other types - the reason probably is easy of booking. For the City hotels, the direct booking rate is slightly higher than the TA/TO rate, not the same for Resort hotel.
```

* Average ADR for cancelled bookings by distribution type for both hotel types:

```{r}
hotels_df %>% select(hotel, is_canceled, distribution_channel, adr) %>%
              filter(is_canceled == 1& adr<400) %>%
              group_by(hotel, is_canceled, distribution_channel) %>%
              summarise(avg_adr = mean(adr)) %>%
              ggplot(., aes(x=distribution_channel, label = round(avg_adr, 0), y=avg_adr, fill=hotel)) +
              geom_segment(aes(x=distribution_channel, xend=distribution_channel, y=0, yend=avg_adr)) +
              geom_point(size=4, alpha=1, shape=21, stroke=0.5) +
              geom_text(size = 3, nudge_y = 1.5, vjust = 2, colour = "blue") +
              theme_light() +
              coord_flip() +
              labs(title ="Average ADR per distribution type", x = "Average ADR",
                       y = "Type of Distribution") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())
```

* Distribution of ADR for cancelled vs not-cancelled:

  + Not-Cancelled

```{r}
library(forcats)
```

```{r}
hotels_df %>% select(hotel, is_canceled, distribution_channel, adr) %>%
              filter(hotel == "City Hotel" & is_canceled == 0 & adr<400) %>%
              ggplot(aes(x=fct_reorder(distribution_channel, adr), y=adr, fill=distribution_channel)) +
              geom_boxplot(alpha=0.7) +
              stat_summary(fun=mean, geom="point", shape=20, size=3, color="red", fill="red") +
              scale_fill_brewer(palette="Set2") +
              labs(title ="Average ADR distribution (non-cancelled) per Hotel Distribution type",
                   x = "Hotel Distribution chanel",
                   y = "Average ADR - log scale") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())
```

```{r}
hotels_df %>% select(hotel, is_canceled, distribution_channel, adr) %>%
              filter(hotel == "City Hotel" & is_canceled == 1& adr<400) %>%
              ggplot(aes(x=fct_reorder(distribution_channel, adr), y=adr, fill=distribution_channel)) +
              geom_boxplot(alpha=0.7) +
              stat_summary(fun=mean, geom="point", shape=20, size=3, color="red", fill="red") +
              scale_fill_brewer(palette="Set2") +
              labs(title ="Average ADR distribution (cancelled) per Hotel Distribution type",
                   x = "Hotel Distribution chanel",
                   y = "Average ADR - log scale") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())

```

```{r}
hotels_df %>% filter(adr > 2000)
```

```{text}
What we can see here is there is a difference between which rate is higher vs lower (Direct-TA/TO) for both types of hotels. For City hotels among cancelled bookings direct booking rate is higher than the TA/TO rate (vice-versa for Resort hotels). If we, compare these findings with the percent cancelled vs not for bot type of hotels (out of total booking) it might be possible that this rate flip plus the deposit taken or not factor had something to do with city hotel's higher percentage of cancellations.

The lollipop graph shows how the rates (high vs low) are flipped for Direct and TA/TO between the City Hotels and the Resort Hotels

Average ADRs for cancelled City Hotels bookings are lower across the board than the non-cancelled bookings.
```

## Looking at who is cancelling more:

* Repeat guest or not:

    + 0 = Not a repeat guest
    + 1 = Repeat guest

* Number of bookings for each guest type:


```{r}
hotels_df %>% select(hotel, is_canceled, is_repeated_guest) %>%
              filter(is_canceled == 0) %>%
              group_by(hotel, is_canceled, is_repeated_guest) %>%
              summarise(bookings_by_guest_type = n()) %>%
              mutate(total_bookings = sum(bookings_by_guest_type)) %>%
              treemap(., 
                      index=c("hotel","is_repeated_guest"),  
                      vSize = "bookings_by_guest_type", 
                      type="index", 
                      palette = "Pastel1",
                      title="Non-cancelled bookings - repeat vs non-repeat guest distribution", 
                      fontsize.title = 14,
                      fontsize.labels=c(15,12),                
                      fontcolor.labels=c("black","blue"),    
                      fontface.labels=c(2,1),                  
                      bg.labels=c("transparent"),              
                      align.labels=list(
                      c("center", "top"), 
                      c("center", "center")),
                      overlap.labels=0.5, inflate.labels=F) 
```
```{r}
hotels_df %>% select(hotel, is_canceled, is_repeated_guest) %>%
              filter(is_canceled == 1) %>%
              group_by(hotel, is_canceled, is_repeated_guest) %>%
              summarise(bookings_by_guest_type = n()) %>%
              mutate(total_bookings = sum(bookings_by_guest_type)) %>%
              treemap(., 
                      index=c("hotel","is_repeated_guest"),  
                      vSize = "bookings_by_guest_type", 
                      type="index", 
                      palette = "Pastel1",
                      title="Cancelled bookings - repeat vs non-repeat guest distribution", 
                      fontsize.title = 14,
                      fontsize.labels=c(15,12),                
                      fontcolor.labels=c("black","blue"),    
                      fontface.labels=c(2,1),                  
                      bg.labels=c("transparent"),              
                      align.labels=list(
                      c("center", "top"), 
                      c("center", "center")),
                      overlap.labels=0.5, inflate.labels=F) 

```
```{r}

hotels_df %>% select(hotel, is_canceled, is_repeated_guest) %>%
              filter(is_canceled == 1) %>%
              group_by(hotel, is_canceled, is_repeated_guest) %>%
              summarise(bookings_by_guest_type = n()) %>%
              mutate(total_bookings = sum(bookings_by_guest_type)) 

```

* Focusing on cancelled bookings:


```{r}
repeat_df = hotels_df %>% select(hotel, is_canceled, is_repeated_guest,
              stays_in_weekend_nights, stays_in_week_nights) %>%
              filter(is_canceled == 1) %>%
              group_by(hotel, is_canceled, is_repeated_guest)%>%
              summarise(bookings_by_guest_type = n(),
              weekend_bookings = sum(stays_in_weekend_nights),
              weekday_bookings = sum(stays_in_week_nights))
```

* Changing binary values to strings for better visual understanding:

```{r}
repeat_df$is_repeated_guest = ifelse(repeat_df$is_repeated_guest == 0,'Not Repeated guest','Repeated Guest')
```

```{r}
repeat_df = repeat_df %>% mutate_if(is.character, as.factor)
```

```{r}
repeat_df
```

```{r}

repeat_df %>% ggplot(., aes(fill=is_repeated_guest, y=log(weekend_bookings),
                            x=hotel, label = weekend_bookings)) +
              geom_bar(position="dodge2", stat="identity") +
              geom_text(size = 3.5, position = position_dodge2(0.9), vjust = 1.8, 
                        colour = "white", fontface = "bold") +
              scale_fill_brewer(palette="Accent")+
              labs(title ="Booking Counts by Repeat/Non-Repeat guests",
                   x = "Type of Hotel",
                   y = "Weekend Bookings (log scale)") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())
              
```

```{r}
repeat_df %>% ggplot(., aes(fill=is_repeated_guest, y=log(weekday_bookings),
                            x=hotel, label = weekday_bookings)) +
              geom_bar(position="dodge2", stat="identity") +
              geom_text(size = 3.5, position = position_dodge2(0.9), vjust = 1.8, 
                        colour = "white", fontface = "bold") +
              scale_fill_brewer(palette="Accent")+
              labs(title ="Booking Counts by Repeat/Non-Repeat guests",
                   x = "Type of Hotel",
                   y = "Weekday Bookings (log scale)") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())
```

```{r}
repeat_df %>% pivot_longer(c(weekend_bookings:weekday_bookings),
              names_to = "Week/Weekend_Booking", values_to="Nights booked")

```
```{r}
repeat_df %>% pivot_longer(c(weekend_bookings:weekday_bookings),
              names_to = "Week_Weekend", values_to="Nights_booked") %>%
              ggplot(., aes(fill=Week_Weekend, y=log(Nights_booked),
                            x=is_repeated_guest, label = Nights_booked)) +
              geom_bar(position="dodge2", stat="identity") + facet_grid(. ~ hotel) +
              geom_text(size = 3, position = position_dodge2(0.9), vjust = 1.8, 
                        colour = "blue", fontface = "bold") +
              scale_fill_brewer(palette="Set3")+
              labs(title ="Booking Counts by Repeat/Non-Repeat guests",
                   x = "Repeat/Non-Repeat Guest",
                   y = "Nights Booked (log scale)") +
              theme(plot.title = element_text(hjust = 0.5), legend.title = element_blank())
  
```


```{text}

Majority of the guests that end up staying are not repeat guests for both hotel types. This same number translates to the majority in cancellations for both hotels. There is a huge difference in numbers for these hotels. Also, both types of guests seem to be staying more during the weekdays. Hotels can use this info to incentivise guests accordingly and retain more of these non-repeat guests as loyal returning guests by incentivising them more. 

```



